---
title: "Gov 50 Final Project"
author: "Nicholas Lopez"
description: "Explaining Florida's Political Trends Over Time"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction
### What factors have primarily contributed to the shift in Florida's voting patterns over time?

In this project, I examine Florida's evolving demographic and socioeconomic conditions over time. In doing so, I attempt to determine which factors are most correlated with the change in Florida's political patterns over time. I study potential factors such as college degree rate, median income, and race. By examining correlations between these trends over the past decades, I attempt to explain shifts in Florida's political status. I hypothesize median income and college degree rate will have the most pronounced effect on voting patterns, which could potentially be confirmed by examining a strong positive correlation between median income and proportion of Republican votes and a strong negative correlation between college degree rate and proportion of Republican votes. 

This is significant for a multitude of reasons. Florida, just a few election cycles ago, was an incredibly significant swing state in US General Elections. Its diversity of backgrounds led to a great diversity of thought and political opinion, and it frequently was the focus of attention on many elections. However, it has since lost this status and is now lauded as one of the most Republican-leaning states, resulting in major victories for the Republican party in recent years. Investigating the factors that are correlated with, and potentially caused, this change would give valuable insight into this change and could potentially be used to identify other states likely to shift allegiances in the coming years. 

## Data
This is an observational study that looks at correlations between several independent variables and one dependent variable over time.

### Data Sources
Our data comes from multiple sources and is thus merged by time. Information on Florida's voting registration comes from the [Florida Department of State](https://dos.fl.gov/elections/data-statistics/voter-registration-statistics/voter-registration-reports/voter-registration-by-party-affiliation/by-party-affiliation-archive/)'s internal data collection methods, where certain information is made available to the public. Data on [income](https://fred.stlouisfed.org/series/MEHOINUSFLA672N) and [education](https://fred.stlouisfed.org/series/GCT1502FL) trends in the state come from Federal Reserve Economic Data, which collects its data from the U.S. Census Bureau. Data on race comes from the [Florida Department of Health](https://www.flhealthcharts.gov/ChartsReports/rdPage.aspx?rdReport=PopAtlas.PopulationAtlasDASHBOARD&rdRequestForwarding=Form), which in turn gets its data from the Florida Legislature, Office of Economic and Demographic Research. It is assumed this data similarly comes from census data. 

### The shift in Florida's voting registration

```{r setup, echo = FALSE}
library(tidyverse)
```

Our dependent variable is the shift in Florida's voting registration over time. In particular, I am tracking the portion of partisan voters in Florida that are registered as Republicans. For the remainder of the project, I will refer to this variable as *repPortion*. I leave out voters not registered as Democrats or Republicans due to increases in third-party registration over time. This allows us to focus on strictly relationships within the two-party system. Below, we see this value graphed across time. 

```{r viz1, echo = FALSE}
registration <- read_csv('data/registration.csv')
registration <- registration |> 
  mutate(repPortion = 100 * republicans/(republicans + democrats))

repPortionPlot <- ggplot(data = registration,
                         mapping = aes(x = year, y = repPortion)) + 
  geom_point() + geom_smooth(method = "loess", se = FALSE, color = 'red') + 
  labs(x='Year', y = 'FL Partisan Voters Registered as Republicans (%)', 
       title = 'Florida Republican Proportion Over Time')
  
repPortionPlot
```

It is this precise trend we wish to explain in this project. 

## Results

### Examining Income
```{r incomePlot, echo = FALSE}
income <- read_csv('data/income.csv')
income <- income |> inner_join(registration)

# scale coefficient between the two variables
coeff <- 1200

repIncomePlot <- ggplot(income, aes(x=year)) +
  
  geom_line( aes(y=repPortion, color = 'Republican Proportion')) + 
  geom_line( aes(y=median_income / coeff, color = 'Median Income')) +
  scale_y_continuous(
    name = "FL Partisan Voters Registered as Republicans (%)",
    sec.axis = sec_axis(~.*coeff, name="Median Income (2022 Real USD)")
  ) + scale_color_manual(values = 
                           c("Republican Proportion" = "red", 
                             "Median Income" = "darkgreen")) + 
  labs(title = 'FL Republican Portion vs. Median Income by Year')

repIncomePlot
```
```{r incomeCorr, echo = FALSE}

income |>
  summarize(cor(median_income, repPortion)) |> 
  knitr::kable(digits = 3, col.names = 
                 c('Correlation between Republican Portion and Median Income by Year in FL'))

```

In the above graph and figure, we see a relatively strong, positive correlation between the (inflation-adjusted) median income in Florida and *repPortion*. This correlation is likely non-trivial. An increased in income is correlated with increases an age, and both of these factors can historically be correlated to increases in likelihood to be registered as a republican. As the median income in Florida rises, it is linked to an increase in suburban neighborhoods, more white-collar positions throughout the state, and a wealthier demographic overall, all of which I believe would drive an increase in republican registration proportions.

A correlation coefficient of 0.685 is moderately strong, which suggests a statistically significant positive correlation. Thus, as the median income increases, we expected the proportion of Florida's partisan voters registered as Republicans to also increase. 

However, we cannot definitively say there is a causal relationship. There are several other confounding variables that would influence these results, from age to college education rates. Thus, we can only conclude there is at least a moderately strong correlation between median income and *repPortion*. 

### Examining Education
```{r educationPlot, echo = FALSE}
education <- read_csv('data/education.csv')
education <- education |> inner_join(registration)

# scale coefficient between the two variables
coeff <- 1

repEducationPlot <- ggplot(education, aes(x=year)) +
  
  geom_line( aes(y=repPortion, color = 'Republican Proportion')) + 
  geom_line( aes(y=degree / coeff, color = '4-Year Degree Rate')) +
  scale_y_continuous(
    name = "FL Partisan Voters Registered as Republicans (%)",
    sec.axis = sec_axis(~.*coeff, name="4-Year Degree Rate (%)")
  ) + scale_color_manual(values = 
                           c("Republican Proportion" = "red", 
                             "4-Year Degree Rate"
                             = "purple")) + 
  labs(title = 'FL Republican Portion vs. 4-Year Degree Rate by Year')

repEducationPlot
```
```{r educationCorr, echo = FALSE}

education |>
  summarize(cor(degree, repPortion)) |> 
  knitr::kable(digits = 3, col.names = 
                 c('Correlation between Republican Portion and Degree by Year in FL'))

```

In the graph above, we see *repPortion* graphed against the 4-Year Degree Rate in Florida, defined as the proportion of residents aged 18 years old and over who have attained a Bachelor's Degree or higher. In the graph, and as represented by the correlation coefficient of 0.496, we see a moderate, positive correlation between the variables. Thus, despite my hypothesis, we do see a positive correlation between the variables.

The explanation for this is likely because of many confounding variables. Over time, in the United States as a whole, 4-year degree rates have increased. Thus, time itself is surely one confounding variable, and there is not necessarily any causal effect between the two variables. 

### Examining Race
```{r racePlot, echo = FALSE}
race <- read_csv('data/race.csv')
race <- race |> mutate(whiteProp = white / total * 100)
race <- race |> inner_join(registration, by = "year")

# scale coefficient between the two variables
coeff <- 1

repRacePlot <- ggplot(race, aes(x=year)) +
  
  geom_line( aes(y=repPortion, color = 'Republican Proportion')) + 
  geom_line( aes(y=whiteProp / coeff, color = 'White Population Proportion')) +
  scale_y_continuous(
    name = "FL Partisan Voters Registered as Republicans (%)",
    sec.axis = sec_axis(~.*coeff, name="White Population Proportion (%)")
  ) + scale_color_manual(values = 
                           c("Republican Proportion" = "red", 
                             "White Population Proportion"
                             = "gray")) + 
  labs(title = 'FL Republican Portion vs. White Population Proportion')

repRacePlot
```
```{r raceCorr, echo = FALSE}

race |>
  summarize(cor(whiteProp, repPortion)) |> 
  knitr::kable(digits = 3, col.names = 
                 c('Correlation between Republican Portion and Degree by Year in FL'))

```

In the graph above, we see *repPortion* graphed against the proportion of residents who identify as white. In the graph, and as represented by the correlation coefficient of -0.76, we see a strong, negative correlation between the variables. Thus, despite my hypothesis, we do see a negative correlation between the variables.

The explanation for this is likely also because of many confounding variables. As the state has developed in the past decades, it has become much more diversified racially, despite its political trends. Thus, time itself is surely one confounding variable, and there is not necessarily any causal effect between the two variables. Furthermore, isolating for exclusively these variables, we would likely see a positive correlation between the two.

## Conclusion

Through this analysis, we have observed a moderately strong correlation between income and the proportion of partisan voters registered as Republicans (*repPortion*) in Florida. This supports the hypothesis. We have observed a moderate correlation between higher education rates and *repPortion*, as well as a strong, negative correlation between the portion of residents who identify as white and *repPortion*. Both of these findings are against the hypothesis. There are certainly many confounding variables within the analysis, such as time, as many of these trends are derived from several other factors. There can be no causal relationship assumed from the data for this reason. With more time, one could examine this data on a county-level scale. Using this, it would be easier to eliminate confounding variables and focus on even more specific ones, such as gentrification in a given area or the presence of certain industries.

## Code

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```

## Sources
[Florida Department of State](https://dos.fl.gov/elections/data-statistics/voter-registration-statistics/voter-registration-reports/voter-registration-by-party-affiliation/by-party-affiliation-archive/)

[Federal Reserve Economic Data, Income](https://fred.stlouisfed.org/series/MEHOINUSFLA672N)

[Federal Reserve Economic Data, Education](https://fred.stlouisfed.org/series/GCT1502FL)

[Florida Department of Health](https://www.flhealthcharts.gov/ChartsReports/rdPage.aspx?rdReport=PopAtlas.PopulationAtlasDASHBOARD&rdRequestForwarding=Form)