---
title: "Gov 50 Final Project"
author: "Nicholas Lopez"
description: "Explaining Florida Election Patterns Over Time"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to changes in voting patterns over time. Specifically, I would be interested in attempting to identify factors (or at least correlations) between changing aspects of a local area (economy, ethnic breakdown, education levels, etc.) and patterns in the party of officers chosen to represent that area (elected congressman, mayors, etc.). This could be done on a local, city-wide scale as well as a state-wide scale. I believe some very interesting patterns and correlations would arise. 

## Proposal

What factors have primarily contributed to the shift in Florida's voting patterns over time? In this project, I plan to study various factors of Florida's demographics and their changes over time. I will attempt to determine which factors are most correlated with the change in Florida's voting patterns over time in US Presidential elections. I will examine the data first on a state level and then attempt to examine data on a county level to find more minute, speicifc demographic shifts. I will look at demographics such as median or average age, college degree rate, median income, race, and possibly other factors. I will then find the correlation between changes in these variables and the proportion of votes for the Republican candidate. I hypothesize age and college degree rate will have the most prounounced effect on voting patterns. This would be confirmed by examining a strong positive correlation between age and proportion of Republican votes and a strong negative correlation between college degree rate and proportion of Republican votes. 

I will be using data from the [Florida Department of State Election Results website](https://results.elections.myflorida.com/). I will download the state and country-level data and merge it with the necessary demographic data. The data is fully comprehensive, and I will attempt to use data dating back to the 1992 election, though the data goes back to 1978. 

## Visualization 

Visualization One:

```{r setup}
library(tidyverse)
```

```{r viz1}
registration <- read_csv('data/registration.csv')
registration <- registration |> mutate(repPortion = 100 * republicans/(republicans + democrats))

repPortionPlot <- ggplot(data = registration,
       mapping = aes(x = year,
                     y = repPortion)) +
  geom_point() + geom_smooth(method = "loess", se = FALSE, color = 'red') + labs(x='Year', y = 'FL Partisan Voters Registered as Republicans (%)')
  
repPortionPlot
```

Graphed, we see the portion of partisan voters in Florida that are registered as Republicans. We leave out voters not registered as Democrats or Republicans due to increases in third-party registration. This allows us to focus on strictly relationships within the two-party system. 

```{r viz2}
registration <- registration |> mutate(demPortion = 100 * democrats/(republicans + democrats))

repPortionPlot <- ggplot(data = registration,
       mapping = aes(x = year,
                     y = demPortion)) +
  geom_point() + geom_smooth(method = "loess", se = FALSE, color = 'blue') + labs(x='Year', y = 'FL Partisan Voters Registered as Democrats (%)')
  
repPortionPlot
```
## Analysis 
```{r analysis1}
income <- read_csv('data/income.csv')
income <- income |> inner_join(registration)

coeff <- 1200

repIncomePlot <- ggplot(income, aes(x=year)) +
  
  geom_line( aes(y=repPortion, color = 'Republican Portion')) + 
  geom_line( aes(y=median_income / coeff, color = 'Median Income')) +
  scale_y_continuous(
    name = "FL Partisan Voters Registered as Republicans (%)",
    sec.axis = sec_axis(~.*coeff, name="Median Income (2022 Real USD)")
  ) + scale_color_manual(values = c("Republican Portion" = "red", "Median Income" = "darkgreen")) + labs(title = 'FL Republican Portion vs. Median Income by Year')

repIncomePlot
```
```{r analysis2}

income |>
  summarize(cor(repPortion, median_income)) |> knitr::kable(digits = 3, col.names = c('Correlation between Republican Portion and Median Income by Year in FL'))
```

In the above graph and figure, we see a relatively strong, positive correlation between the (inflation-adjusted) median income in FL and the portion of partisan voters in the state registered as Republicans. This correlation is likely non-trivial, and I would assume there is a decent relationship between other variables to be explored in the remainder of the project, such as median age across the state. An increased in income is correlated with increases an age, and both of these factors can historically be correlated to increases in likelihood to be registered as a republican. Across the past 50 years in Florida, I hypothesize we will begin to see similar patterns in the demographic makeup and correlations within the state. As the median income in Florida rises, it is linked to an increase in suburban neighborhoods, more white-collar positions throughout the state, and a wealthier demographic overall, all of which I believe would drive an increase in republican registration proportions.

## Sources
[USA Facts](https://usafacts.org/data/topics/people-society/population-and-demographics/our-changing-population/state/florida/?endDate=2021-01-01&startDate=1972-01-01)

[Florida Department of State](https://dos.fl.gov/elections/data-statistics/voter-registration-statistics/voter-registration-reports/voter-registration-by-party-affiliation/by-party-affiliation-archive/)

[Federal Reserve Bank Of St. Louis](https://fred.stlouisfed.org/series/MEHOINUSFLA672N)

[Statista](https://www.statista.com/statistics/205798/median-household-income-in-florida/)
