---
title: "Gov 50 Final Project"
author: "Nicholas Lopez"
description: "Explaining Florida Election Patterns Over Time"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to changes in voting patterns over time. Specifically, I would be interested in attempting to identify factors (or at least correlations) between changing aspects of a local area (economy, ethnic breakdown, education levels, etc.) and patterns in the party of officers chosen to represent that area (elected congressman, mayors, etc.). This could be done on a local, city-wide scale as well as a state-wide scale. I believe some very interesting patterns and correlations would arise. 

## Proposal

What factors have primarily contributed to the shift in Florida's voting patterns over time? In this project, I plan to study various factors of Florida's demographics and their changes over time. I will attempt to determine which factors are most correlated with the change in Florida's voting patterns over time in US Presidential elections. I will examine the data first on a state level and then attempt to examine data on a county level to find more minute, speicifc demographic shifts. I will look at demographics such as median or average age, college degree rate, median income, race, and possibly other factors. I will then find the correlation between changes in these variables and the proportion of votes for the Republican candidate. I hypothesize age and college degree rate will have the most prounounced effect on voting patterns. This would be confirmed by examining a strong positive correlation between age and proportion of Republican votes and a strong negative correlation between college degree rate and proportion of Republican votes. 

I will be using data from the [Florida Department of State Election Results website](https://results.elections.myflorida.com/). I will download the state and country-level data and merge it with the necessary demographic data. The data is fully comprehensive, and I will attempt to use data dating back to the 1992 election, though the data goes back to 1978. 

Visualization One:

```{r setup}
library(tidyverse)
```

```{r viz1}
registration <- read_csv('data/registration.csv')
registration <- registration |> mutate(repPortion = 100 * republicans/(republicans + democrats))

repPortionPlot <- ggplot(data = registration,
       mapping = aes(x = year,
                     y = repPortion)) +
  geom_point() + geom_smooth(method = "loess", se = FALSE, color = 'red') + labs(x='Year', y = 'Partisan Voters Registered as Republicans (%)')
  
repPortionPlot
```

Graphed, we see the portion of partisan voters in Florida that are registered as Republicans. We leave out voters not registered as Democrats or Republicans due to increases in third-party registration. This allows us to focus on strictly relationships within the two-party system. 

```{r viz2}
registration <- registration |> mutate(demPortion = 100 * democrats/(republicans + democrats))

repPortionPlot <- ggplot(data = registration,
       mapping = aes(x = year,
                     y = demPortion)) +
  geom_point() + geom_smooth(method = "loess", se = FALSE, color = 'blue') + labs(x='Year', y = 'Partisan Voters Registered as Democrats (%)')
  
repPortionPlot
```

